---
title: "lab_12"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   tidytext and our usual libraries

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
library(tidyverse)
library(tidytext)
library(janitor)
library(lubridate)
library(rvest)
```

## Questions

**Q1.** You've been assigned to report a story about the leading reasons that Maryland attorneys get sanctioned by the state for misconduct. The state [publishes lists of sanctions](https://www.courts.state.md.us/attygrievance/sanctions) that contain a short text description about the situation. Load the CSV file in the data folder containing records from fiscal year 2011 onwards. Make a list of unique words from the text column, then following the example in the pre_lab, remove common "stop words" from that list and create a list of the top 10 words containing the percentage of occurrences each word represents. What's the leading word in that answer and, broadly, what do you think the top 10 words describe?

**The leading word in this answer is "failing," overall, the top 10 words mainly discusses basic law terms such as attorney, trust and client. However, there are some words that have negative connotations to them such as failed, as well as words dealing with accounts and conduct.**

```{r}
md_attorney_words <- read_csv("data/md_attorney_sanctions.csv") 

md_unique_words <- md_attorney_words %>% select(text) %>% unnest_tokens(word, text)

data("stop_words")

md_unique_words %>% 
  anti_join(stop_words) %>% 
  group_by(word) %>%
  tally(sort = TRUE) %>%
  mutate(percent = (n/sum(n)*100)) %>%
  top_n(10)
```

**Q2.** Let's move beyond single words to phrases. Make a list of the top 10 three-word phrases, called trigrams, based on the example from the pre_lab (you'll need to modify the example code to do this). What's the top trigram and how often does it appear? What does that phrase mean in legal terms?

**The top trigram is "the right to," this phrase means in legal terms that an individual has a right to certian things, either provided by via a lawyer, or through judicial process and through the courts.**

```{r}
md_attorney_words %>%
  unnest_tokens(trigram, text, token = "ngrams",n=3) %>%
  separate(trigram,c("word1", "word2", "word3"), sep = " ") %>%
  filter(word1 %in% stop_words$word) %>%
  filter(word2 %in% stop_words$word) %>%
  filter(word3 %in% stop_words$word) %>%
  mutate(trigram = paste(word1,word2,word3, sep = " ")) %>%
  group_by(trigram) %>%
  tally(sort = TRUE) 
```

**Q3.** Let's drop back down to more traditional text analysis - take the top trigram from Q2 and write code to see how many times it occurs in the text column in each fiscal year. What do you think the answer produced by your code suggests? What else could you do to try and clarify the most important reasons attorneys get sanctioned?

**The answer by the code suggests that it has been used more from the mid, to end of the 2010s, as compared to 2020 and the beginning of the 2010s. Something else I could try to put more negative connotation-like words to see if I can find words with negative connotations that were used the most.**

```{r}

trigram_counts <- md_attorney_words %>%
  mutate(trigram_present = str_count(text, "the right to"))

trigram_by_year <- trigram_counts %>%
  group_by(fiscal_year) %>%
  summarise(total_trigram_count = sum(trigram_present, na.rm = TRUE))

trigram_by_year

```
