---
title: "Reverse Engineering Project Group 2"
author: "Lillian Glaros, William Deffaa, Jake Tiger"
date: "10/22/2025"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this notebook, we are reverse engineering the story, [Thousands of kids are missing from school. Where did they go?](https://projects.apnews.com/features/2023/missing-children/index.html)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data.

```{r}
# Load required data
public_enrollment <- read_csv("data/public-enrollment-by-grade.csv")
private_enrollment <- read_csv("data/private-enrollment-by-grade.csv")
pop_by_grade <- read_csv("data/population-by-age.csv")

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 
cleaned_private_enrollment <- private_enrollment |>
  mutate(grade = str_to_upper(grade))

cleaned_public_enrollment <- public_enrollment |>
  mutate(grade = str_to_upper(grade))
```

## Sentences to Engineer

In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

-   **Sentence text**: Public school enrollment plunged during the pandemic. Our analysis shows where the students went. In the 21 states where data was available, enrollment fell by **about 700,000 students** between the 2019-2020 and 2021-2022 school years.
-   **Analysis summary**: We did the same analysis as the AP story, but the results were much different because the data we're using is more complete. The AP story found that public-school enrollment declined by about 700,000 students, but our analysis showed a decrease of over 2.4 million students due to having data from more states. We summed up enrollment totals for each year and found the difference between the 2019-2020 and 2021-2022 school years.

```{r}
cleaned_public_enrollment |>
  
  group_by(sch_year) |>
  summarise(total_enrollment = sum(enrollment, na.rm=TRUE)) |>
  
  
arrange(desc(total_enrollment))
```

```{r}
98405826 - 95994012
```

### Sentence 2

-   **Sentence text**: Some of those students enrolled in private schools. Across the states in the analysis, private-school enrollment grew by **about 103,000 students**.
-   **Analysis summary**:

```{r}
cleaned_private_enrollment |>
  
  group_by(sch_year) |>
  summarise(total_enrollment = sum(enrollment, na.rm=TRUE)) |>
  
  
arrange(desc(total_enrollment))


```

```{r}

```

### Sentence 3

-   **Sentence text**: Some families also moved out of state, accounting for some enrollment decline. And birth rates in some states are falling. The population of school-age kids in the states we studied fell by **about 183,000**, according to census estimates.
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
dates_pop_grade <- pop_by_grade %>%
  filter(str_detect(est_date,"2020-07") | str_detect(est_date, "2021-07"))

dates_pop_grade %>% 
  group_by(est_date) %>%
  summarise(
    total_population = sum(population)
    )


```

### Sentence 4

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-
