---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse

# Load janitor
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

# Load in the Maryland grants and loan data
```

## Answer questions

**Q1:** You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

**A1: After narrowing it down, I was able to find 55 records that referenced Prince George's County Public Schools.**

```{r}
pg_schools <- payments %>%
  filter(str_detect(`Vendor Name`, "PRINCE GEORGES COUNTY PUBLIC SCHOOLS") | str_detect(`Vendor Name`,"PG COUNTY PUBLIC SCHOOL") |
str_detect(`Vendor Name`,"PRINCE GEORGES PUBLIC SCHOOLS") |
str_detect(`Vendor Name`,"PRINCE GEORGES COUNTY SCHOOL") |
str_detect(`Vendor Name`,"PRINCE GEORGES COUNTY PUBLIC SCHOOL") |
str_detect(`Vendor Name`,"PRINCE GEORGE'S COUNTY PUBLIC SCHOOLS")) %>%
  filter(str_detect(Category, "Vendor Payment")) %>%
  arrange(desc(`Fiscal Year`))

```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

**A2:** **The fiscal year that had the largest total amount was 2019 and the aspect that stood out the most to me was the fact that 2014 was second to last, while 2015 jumped up to second in terms of total amount.**

```{r}
pg_schools %>%
  group_by(`Fiscal Year`) %>%
  summarise(total_payments_yearly = sum(Amount)) %>%
  arrange(desc(total_payments_yearly))
  
  
 
  
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

**A3.** **Some cleaning that would need to be done would make sure I would only have one row per vendor would be to clean up all the different names of vendors by putting them into a singular name such as "UNIVERSITY OF MARYLAND COLLEGE PARK."**

```{r}
umd_grants_loans <- cleaned_md_grants_loans %>%
  filter(str_detect(grantee, "THE UNIVERSITY OF MARYLAND") | 
           str_detect(grantee, "UNIVERSITY OF MARYLAND") |
           str_detect(grantee, "UNIVERSITY OF MARYLAND COLLEGE PARK") |
           str_detect(grantee, "UMD") | str_detect(zip5, "20740") | str_detect(grantee, "UNIVERSITY OF MD COLLEGE PARK")) 
 

```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

**A4: Stem cell research funding has mostly fluctuated over time, aside from 2009 and 2023 (when it started and when the most recent year recorded was), where there were extremely large amounts of funding being put into stem cell research.**

```{r}
cleaned_md_grants_loans %>%
  filter(str_detect(description, "STEM CELL RESEARCH GRANT")) %>%
  group_by(fiscal_year) %>%
  summarise(total_payments_fiscal = sum(amount)) %>%
  arrange(desc(total_payments_fiscal))

```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

**A5: There are 30 grantees within the College Park, MD zip code (20740). Something I would need to do to clean the results so I only have one record per grantee would be to change recurring names of the same meaning or grantee, into a singular name that can be universally used. An example would be to use the name "UNIVERSITY OF MARYLAND COLLEGE PARK" for any record that has the University of Maryland (or MD) in the name.**

```{r}
cleaned_md_grants_loans %>%
  filter(str_detect(zip5, "20740")) %>%
  group_by(grantee) %>%
  summarise(total_payments = sum(amount)) %>%
  arrange(desc(total_payments))
```
