---
title: "lab_01"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you?

**The School of Music is the most surprising one to me because I don't think I have honestly met anyone within the school of music, I never realized that it is a popular major at UMD.**

```{r}
umd_courses %>% 
  group_by(department) %>% 
  summarise(
    mean_seats = n()
    ) %>%
    arrange(desc(mean_seats))
```

**Q2** Find out the greatest number of times any individual class has been offered in our data set. Your code must show the greatest number of classes first.

**Doctoral Dissertation Research seems to be the most offered class throughout time in our data set.**

```{r}
umd_courses %>% 
  group_by(title) %>%
  summarise(
    count_classes = n()
  ) %>% 
  arrange(desc(count_classes))
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most? **The instructor that has taught the most is Jason Kuo.**

```{r}
umd_courses %>% 
  group_by(instructors) %>%
  summarise(
    count_classes = n()
  ) %>% 
  arrange(desc(count_classes))
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why?

**I think that what would make the answer easier to see would be to be able to have the corresponding department next to the person who has taught the most classes. Unfortunately, I wan't able to do that and could only get the instructor who has taught the most. However, I am still very confident I got the right answer.**

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting?

**The question I tried to answer was "Has there been a change in the amount of seats being offered per department over time, say by semester?" I don't think I was fully able to answer the question because I wanted to have the terms in a more readable format so people could understand what terms are which.**

```{r}
umd_courses %>% 
  group_by(term, department) %>%
  summarise(
    total_seats = sum(seats)
    ) %>% 
  arrange(desc(total_seats))

```
